# Energy demand forecast

## Synopsis

The objective of this analysis is to develop a mathematical model for energy demand forecasting in Spain. The model should be able to predict tomorrow's demand having data of the daily demand the days before. The research is done using the R programming language and all the data and code needed to reproduce the study is provided within this document to ensure fully reproducible research.

The data is available here:

```{r, cache = TRUE}
dataurl <- 'https://gist.githubusercontent.com/Peque/715e91350f0e68e3342f/raw/d28312ac0e49888a5079fcea188770acaf3aa4a2/mme.csv'
```

It is a CSV file with values of the daily energy demand (in GWh) between 2004 and 2012. This data has been divided in two sets:

* A set to create the model, containing the data between 20014 and 2010.
* A set to calculate the forecasting error, with the data of 2011 and 2012.

The source code of this $Rmd$ file is available on Github:

https://github.com/Peque/peque/blob/master/etsii/mme/mme.Rmd

The following extra packages have been used for the analysis:

```{r, message = FALSE}
library(xts)
library(ggplot2)
library(forecast)
```

## Exploratory data analysis

Downloading the data:

```{r, cache = TRUE, fig.width = 10}
# Downloading and loading data into memory
tmp <- tempfile()
download.file(dataurl, tmp, method = 'curl')
df <- read.csv(tmp)
unlink(tmp)
```

For later use, the dataframe is formatted and completed with extra information gotten from the raw data. Dates are converted to POSIX format and new columns are added containing information about the day of the week and the day of the year.

```{r, cache = TRUE}
# Convert date strings to POSIX dates
df$date <- strptime(df$date, format = '%d-%m-%y')
# Day of the week
df$day <- as.factor(strftime(df$date, format = '%A'))
# Day of the year
df$yearday <- as.factor(strftime(df$date, format = '%m%d'))
# Final structure for the study
str(df)
```

The data set is divided to create a test set:

```{r}
df_test <- subset(df, date >= strptime('01-01-2011', format = '%d-%m-%Y'))
df <- subset(df, date < strptime('01-01-2011', format = '%d-%m-%Y'))
ts <- ts(df$demand, frequency = 1)
```

A time series plot helps visualizing the demand evolution across the time period.

```{r, cache = TRUE, fig.width = 10}
# Dataframe and time series objects
demandts <- xts(df$demand, df$date)
plot(demandts, main = 'Energy demand evolution', xlab = 'Date', ylab = 'Demand (GWh)')
```

A seasonal dependency of the demand can be easily spotted in the graphics, although there are other factors that may affect the results, such as the temperature, holidays, weekends, etc..

During weekends, the demand decreases considerably compared to the rest of the week days.

```{r, cache = TRUE, fig.width = 10}
# Demand by day of the week
ggplot(df, aes(day, demand)) + geom_boxplot() + xlab('Day') + ylab('Demand (GWh)') + ggtitle('Demand per day of the week')
```

During winter and summer the demand is clearly higher exept for, probably, vacation periods. Holydays are also easily spotted in the graphics, being the lowest peaks of demand.

```{r, fig.width = 10}
# Aggregating demand by day of the year (average)
avg_demand_per_yearday <- aggregate(demand ~ yearday, df, 'mean')

# Computing the smooth curve for the time series. Data is replicated before computing the curve in order to achieve continuity
smooth_yearday <- rbind(avg_demand_per_yearday, avg_demand_per_yearday, avg_demand_per_yearday, avg_demand_per_yearday, avg_demand_per_yearday)
smooth_yearday <- lowess(smooth_yearday$demand, f = 1 / 45)
l <- length(avg_demand_per_yearday$demand)
l0 <- 2 * l + 1
l1 <- 3 * l
smooth_yearday <- smooth_yearday$y[l0:l1]

# Plotting the result
par(mfrow = c(1, 1))
# Setting year to 2000 to allow existence of 29th February
dates <- as.Date(paste(levels(df$yearday), '2000'), format = '%m%d%Y')
plot(dates, avg_demand_per_yearday$demand, type = 'l', main = 'Average daily demand', xlab = 'Date', ylab = 'Demand (GWh)')
lines(dates, smooth_yearday, col = 'blue', lwd = 2)
```

The graphics bellow show the errors. Notice how the biggest errors are all negative.

```{r, fig.width = 10}
par(mfrow = c(1, 2))
diff <- avg_demand_per_yearday$demand - smooth_yearday
abs_diff <- abs(diff)
barplot(diff[order(-abs_diff)], main = 'Smoothing error', ylab = 'Error')
boxplot(diff, main = 'Smoothing error', ylab = 'Error')
```

The exact dates which are generating those errors are, indeed, holidays or the day just before holidays (as is the case for the 25th November and 31th Devember).

```{r}
head(strftime(dates[order(-abs_diff)], format = '%B %d'), 10)
```

The autocorrelation function shows a highly autocorrelated seasonal non-stationary process with, as expected, yearly and weekly cicles. The ACF alone, however, tells us little about the orders of dependence for ARMA or AR processes. The PACF is better for AR models, and also shows the weekly and yearly seasons, although the correlation is lost faster with the lag.

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(2, 2))
acf(df$demand, 100, main = 'Autocorrelation')
acf(df$demand, 1500, main = 'Autocorrelation')
pacf(df$demand, 100, main = 'Partial autocorrelation')
pacf(df$demand, 1500, main = 'Partial autocorrelation')
```

## Further analysis

Decomposition of the weekly seasonal time series.

```{r, fig.width = 10, fig.height = 10}
wts <- ts(ts, frequency = 7)
dec_wts <- decompose(wts)
plot(dec_wts)
# Demand minus week seasonal
df$demand_mws <- df$demand - as.numeric(dec_wts$season)
```

Decomposition of the yearly seasonal time series. 29th February days are excluded for frequency matching. The time series is formed out of the original observation minus the weekly seasonal data.

```{r, fig.width = 10, fig.height = 10}
yts <- ts(subset(df, yearday != '0229')$demand_mws, frequency = 365)
dec_yts <- decompose(yts)
plot(dec_yts)
days365 <- which(df$yearday != '0229')
february29ths <- which(df$yearday == '0229')
df$demand_mwys[days365] <- df$demand_mws[days365] - as.numeric(dec_yts$season)
# Fill values on February 29th
df$demand_mwys[february29ths] <- df$demand_mws[february29ths]
```

A new time series is formed out of the original observation minus the weekly and the yearly seasonal data:

```{r, cache = TRUE, fig.width = 10}
par(mfrow = c(1, 1))
ts_mwys <- ts(df$demand_mwys, frequency = 1)
demandts_mwys <- xts(df$demand_mwys, df$date)
plot(demandts_mwys, main = 'Energy demand minus seasonal data', xlab = 'Date', ylab = 'Demand (GWh)')
```

Plotting the average daily demand of the demand minus the seasonal data shows a new error rate much lower than the one seen before:

```{r, fig.width = 10}
# Aggregating demand by day of the year (average)
avg_demand_mwys_per_yearday <- aggregate(demand_mwys ~ yearday, df, 'mean')

# Computing the smooth curve for the time series. Data is replicated before computing the curve in order to achieve continuity
smooth_yearday <- rbind(avg_demand_mwys_per_yearday, avg_demand_mwys_per_yearday, avg_demand_mwys_per_yearday, avg_demand_mwys_per_yearday, avg_demand_mwys_per_yearday)
smooth_yearday <- lowess(smooth_yearday$demand_mwys, f = 1 / 45)
l <- length(avg_demand_mwys_per_yearday$demand_mwys)
l0 <- 2 * l + 1
l1 <- 3 * l
smooth_yearday <- smooth_yearday$y[l0:l1]

# Plotting the result
par(mfrow = c(1, 1))
# Setting year to 2000 to allow existence of 29th February
dates <- as.Date(paste(levels(df$yearday), '2000'), format = '%m%d%Y')
plot(dates, avg_demand_mwys_per_yearday$demand_mwys, type = 'l', main = 'Average daily demand', xlab = 'Date', ylab = 'Demand (GWh)')
lines(dates, smooth_yearday, col = 'blue', lwd = 2)
```

New errors:

```{r, fig.width = 10}
par(mfrow = c(1, 2))
diff <- avg_demand_mwys_per_yearday$demand_mwys - smooth_yearday
abs_diff <- abs(diff)
barplot(diff[order(-abs_diff)], main = 'Smoothing error', ylab = 'Error')
boxplot(diff, main = 'Smoothing error', ylab = 'Error')
```

The new ACF and PACF are as follow now:

```{r, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))
acf(df$demand_mwys, 100, main = 'Autocorrelation')
pacf(df$demand_mwys, 100, main = 'Partial autocorrelation')
```

## SARIMA model

The initial ARIMA parameters have been found using the R $auto.arima()$ function. The differencing parameter $d$ is selected using the KPSS test. If the null hypothesis of stationarity is accepted when the KPSS is applied to the original time series, then $d = 0$. Otherwise, the series is differenced until the KPSS accepts the null hypothesis. After that, $p$ and $q$ are selected using either AIC or BIC. The SARIMA model has been created using those ARIMA parameters.

```{r, cache = TRUE}
model <- Arima(ts, order = c(2, 1, 2), list(order = c(1, 1, 1), period = 7))
```

Forecasting error can be calculated iterating through the test data frame:

```{r, cache = TRUE, fig.width = 10}
auxts <- ts
auxmodel <- model
errs <- c()
pred <- c()
perc <- c()
for (i in 1:nrow(df_test)) {
  p <- as.numeric(predict(auxmodel, newdata = auxts, n.ahead = 1)$pred)
  pred <- c(pred, p)
  errs <- c(errs, p - df_test$demand[i])
  perc <- c(perc, (p - df_test$demand[i]) / df_test$demand[i])
  auxts <- ts(c(auxts, df_test$demand[i]), frequency = 7)
  auxmodel <- Arima(auxts, model = auxmodel)
}
par(mfrow = c(1, 1))
plot(errs, type = 'l', main = 'Error in the forecast')
plot(pred, type = 'l', main = 'Real vs. forecast', col = 'red')
lines(df_test$demand)
legend('topright', c('Real', 'Forecast'), lty = 1, col = c('black', 'red'))
abserr <- mean(abs(errs))
percerr <- mean(abs(perc)) * 100
```

The mean error across the test data frame is `r abserr` GWh (**`r percerr`%**).

As it was shown before, some special days present less demand than others. Those days me be taken into account in order to reduce the error:

```{r, cache = TRUE, fig.width = 10}
specialday <- function(day) {
  correction = 0
  if (format(day, '%m%d') %in% c('0101', '0501', '0106', '0815', '1012', '1101', '1206', '1208', '1224', '1225', '1226', '1231'))
      correction = -100
  else if (format(day, '%m%d') %in% c('0319'))
    correction = -50
  # On Sunday, do not apply correction
  if (as.factor(strftime(day, format = '%A')) == 'Sunday')
    return(0)
  return(correction)
}

model <- Arima(ts, order = c(2, 1, 2), list(order = c(1, 1, 1), period = 7))
auxts <- ts
auxmodel <- model
errs <- c()
pred <- c()
perc <- c()
for (i in 1:nrow(df_test)) {
  p <- as.numeric(predict(auxmodel, newdata = auxts, n.ahead = 1)$pred)
  correction = specialday(df_test$date[i])
  pred <- c(pred, p + correction)
  errs <- c(errs, p + correction - df_test$demand[i])
  perc <- c(perc, (p + correction - df_test$demand[i]) / df_test$demand[i])
  if (!correction)
    auxts <- ts(c(auxts, df_test$demand[i]), frequency = 7)
  else
    auxts <- ts(c(auxts, p), frequency = 7)
  auxmodel <- Arima(auxts, model = auxmodel)
}
par(mfrow = c(1, 1))
plot(errs, type = 'l', main = 'Error in the forecast')
plot(pred, type = 'l', main = 'Real vs. forecast', col = 'red')
lines(df_test$demand)
legend('topright', c('Real', 'Forecast'), lty = 1, col = c('black', 'red'))
abserr <- mean(abs(errs))
percerr <- mean(abs(perc)) * 100
```

The new mean error across the test data frame is `r abserr` GWh (**`r percerr`%**).

## Forecast

An example of the model's forecast capabilities:

```{r, fig.width = 10}
plot(forecast(Arima(tail(ts, 200), model = model)))
```

## References

- R. H. Shumway, D. S. Stoffer. *Time Series Analysis and Its Applications*. 2010.
- R. J. Hyndman. *Forecasting: principles and practice*. 2013.
- P. S. P. Cowpertwait, A. V. Metcalfe. *Introductory Time Series with R*. 2009.
